---
title: "Data 606 Project Proposal"
author: "David Blumenstiel"
date: "3/28/2020"
output:
  pdf_document: default
  html_document: default
---

### Data Preparation

```{r setup, echo=TRUE, results='hide', warning=FALSE, message=FALSE}
#Loading the dataframe
df <- read.csv('https://raw.githubusercontent.com/davidblumenstiel/data/master/kaglemovies/tmdb_5000_movies.csv')

#Trimming out the parts variables we aren't interested in
movies <- df[,c(1,9,13,14,19)]

#Taking out observations which have a budget under 1000 (gets rid of some suspect observations)
movies <- movies[movies$budget > 1000,]

```


### Research question 

I want to determine what relationship the budget, popularity, runtime, and average rating of a movie have on it's revenue


### Cases 

The cases are individual movies that were previously releasesd (last updated September 2017) which had a budget over $1000; there are 3734 cases.


### Data collection 

The authors of the dataset collected the data from https://www.themoviedb.org using their API.  I obtained the dataset from https://www.kaggle.com/tmdb/tmdb-movie-metadata/discussion.  I also cleaned it a bit by taking out the variables that I will not be looking at, and observations with budgets under $1000 to reduce what I think may be unrealistic observations.


### Type of study 

This is an observational study.


### Data Source 

Data was collected by The Movie Database (TMDb) https://www.themoviedb.org, and presented on Kagle https://www.kaggle.com/tmdb/tmdb-movie-metadata.



### Dependent Variable

The responce variable is revenue, and it is numerical



### Independent Variable

The explanatory variables are budget, popularity, runtime, and average rating.


### Relevant summary statistics 

```{r}
#Revenue
summary(movies$revenue)
hist(movies$revenue)


#Budget
summary(movies$budget)
hist(movies$budget)


#Popularity
summary(movies$popularity)
hist(movies$popularity)


#Runtime
summary(movies$runtime)
hist(movies$runtime)


#Average Rating
summary(movies$vote_average)
hist(movies$vote_average)


```