---
title: "Chapter 1 - Introduction to Data"
author: ''
output:
  html_document:
    df_print: paged
  pdf_document:
    extra_dependencies:
    - geometry
    - multicol
    - multirow
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Smoking habits of UK residents**. (1.10, p. 20) A survey was conducted to study the smoking habits of UK residents. Below is a data matrix displaying a portion of the data collected in this survey. Note that "$\pounds$" stands for British Pounds Sterling, "cig" stands for cigarettes, and "N/A" refers to a missing component of the data.

\begin{center}
\scriptsize{
\begin{tabular}{rccccccc}
\hline
	& sex 	 & age 	& marital 	& grossIncome 					     & smoke & amtWeekends	& amtWeekdays \\ 
\hline
1 	& Female & 42 	& Single 	& Under $\pounds$2,600 			     & Yes 	 & 12 cig/day   & 12 cig/day \\ 
2 	& Male	 & 44	& Single 	& $\pounds$10,400 to $\pounds$15,600 & No	 & N/A 			& N/A \\ 
3 	& Male 	 & 53 	& Married   & Above $\pounds$36,400 		     & Yes 	 & 6 cig/day 	& 6 cig/day \\ 
\vdots & \vdots & \vdots & \vdots & \vdots 				             & \vdots & \vdots 	    & \vdots \\ 
1691 & Male  & 40   & Single 	& $\pounds$2,600 to $\pounds$5,200   & Yes 	 & 8 cig/day 	& 8 cig/day \\   
\hline
\end{tabular}
}
\end{center}


(a) What does each row of the data matrix represent?

    *Each row represents a person in the study*

(b) How many participants were included in the survey?

    *1691*

(c) Indicate whether each variable in the study is numerical or categorical. If numerical, identify as continuous or discrete. If categorical, indicate if the variable is ordinal.

    *sex:categorical, age:numerical-continual, marital:categorical, grossIncome:categorical-ordinal, smoke:categorical, amtWeekends:numerical-continual,  amtWeekdays:numerical-continual*

```{r, eval=FALSE, echo=FALSE}
# NOTE: It is not absolutely necessary to load the data to answer these 
# questions, but is here for further exploration. 
if(!file.exists('11263-Smoking_tcm86-13253.xls')) {
	download.file('https://github.com/jbryer/DATA606Fall2019/blob/master/course_data/11263-Smoking_tcm86-13253.xls?raw=true', 
				  destfile = '11263-Smoking_tcm86-13253.xls')
}
smoking <- readxl::read_excel('11263-Smoking_tcm86-13253.xls')
```





--------------------------------------------------------------------------------

\clearpage

**Cheaters, scope of inference**. (1.14, p. 29) Exercise 1.5 introduces a study where researchers studying the relationship between honesty, age, and self-control conducted an experiment on 160 children between the ages of 5 and 15[^1]. The researchers asked each child to toss a fair coin in private and to record the outcome (white or black) on a paper sheet, and said they would only reward children who report white. Half the students were explicitly told not to cheat and the others were not given any explicit instructions. Differences were observed in the cheating rates in the instruction and no instruction groups, as well as some differences across children's characteristics within each group.

(a) Identify the population of interest and the sample in this study.

    *The population of interest include children between 5 and 15 years, and the sample are the 160 children.*
    
(b) Comment on whether or not the results of the study can be generalized to the population, and
    
    *if the findings of the study can be used to establish causal relationships.  If the children were selected at random, then you could generalize the results (specific to coin tossing) to the population. The study can be used to establish casual relationships within the population.*

[^1]: Alessandro Bucciol and Marco Piovesan. "Luck or cheating? A field experiment on honesty with children". In: Journal of Economic Psychology 32.1 (2011), pp. 73-78. Available at https://papers.ssrn.com/sol3/papers.cfm?abstract_id=1307694



--------------------------------------------------------------------------------

\clearpage

**Reading the paper**. (1.28, p. 31) Below are excerpts from two articles published in the NY Times:

(a) An article titled Risks: Smokers Found More Prone to Dementia states the following:

"Researchers analyzed data from 23,123 health plan members who participated in a voluntary exam and health behavior survey from 1978 to 1985, when they were 50-60 years old. 23 years later, about 25% of the group had dementia, including 1,136 with Alzheimer's disease and 416 with vascular dementia. After adjusting for other factors, the researchers concluded that pack-a- day smokers were 37% more likely than nonsmokers to develop dementia, and the risks went up with increased smoking; 44% for one to two packs a day; and twice the risk for more than two packs."

Based on this study, can we conclude that smoking causes dementia later in life? Explain your reasoning.

*We can't conclude that smoking causes dementia.  There other things that may affect whether or not someone develops dementia other than smoking, which may or may not have been controled for in this study (it wasn't very specific).  We can however say that people who were smokers at 50-60 years old are at a greater risk of developing dementia down the road.*


(b) Another article titled The School Bully Is Sleepy states the following:

"The University of Michigan study, collected survey data from parents on each child's sleep habits and asked both parents and teachers to assess behavioral concerns. About a third of the students studied were identified by parents or teachers as having problems with disruptive behavior or bullying. The researchers found that children who had behavioral issues and those who were identified as bullies were twice as likely to have shown symptoms of sleep disorders."

A friend of yours who read the article says, "The study shows that sleep disorders lead to bullying in school children." Is this statement justified? If not, how best can you describe the conclusion that can be drawn from this study?

*We cannot conclude that sleep disorders lead to bullying.  First, the study mentiones that both children who have behavioral issues (which could be anything) and bullies as a group.  If sleep disorders did have an effect, it could be it only affects other behavoiral issues and not bullying.  Second, if there was a causal relationship, whose to say it's not bullying that causes sleep disorderes and not the other way around?  Third, there are likely confounding factors involved.  E.g., maybe bullies have sleep disorders only becuase they have chaotic households, which itself could lead to bullying behavior.*





--------------------------------------------------------------------------------

\clearpage

**Exercise and mental health.** (1.34, p. 35) A researcher is interested in the effects of exercise on mental health and he proposes the following study: Use stratified random sampling to ensure rep- resentative proportions of 18-30, 31-40 and 41-55 year olds from the population. Next, randomly assign half the subjects from each age group to exercise twice a week, and instruct the rest not to exercise. Conduct a mental health exam at the beginning and at the end of the study, and compare the results.

(a) What type of study is this?
    
    *This is an experemental study*
    
(b) What are the treatment and control groups in this study?
   
    *The treatment group is the one perscribed exercise twice and week.  I think the control group is the one instructed not to exercise, but that sounds more like a second treatment to me.*
    
(c) Does this study make use of blocking? If so, what is the blocking variable?
    
    *It does by age group*

(d) Does this study make use of blinding?
   
    *No*
    
(e) Comment on whether or not the results of the study can be used to establish a causal rela- tionship between exercise and mental health, and indicate whether or not the conclusions can be generalized to the population at large.  

    *This is an experemental study, and is thus in a position to provide strong evidence as to any causal relationship.  The results are most applicable to their respective block ages (it may not apply to those under 18 or obver 55).*
    
(f) Suppose you are given the task of determining if this proposed study should get funding. Would you have any reservations about the study proposal?
   
    *The study doesn't really have a control, so much as it does two treatments (exercise or no exercise).  A better study may involve randomly putting participants into three groups, telling one to exercise, one to not, and not telling the third to change anything.  Some form of blinding (say, not telling the subjects about the purpose of the study or the other groups) would also be good.*



